Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation 

### 2025-04-08 manuscript/index.qmd
improved introduction

---
title: "Introduction for a BART-Based Model to Predict Chlorophyll-a from Satellite Reflectance"
author: Erdem Karaköylü
date: "2025-04-03"
format:
  pdf:
    documentclass: agujournal2019
    template: templates/agujournaltemplate.tex
    keep-tex: true
bibliography: references.bib
---

@@

#### Historical Context of Chlorophyll Algorithms
Satellite ocean color missions have long relied on empirical spectral algorithms to estimate near-surface chlorophyll-a (Chl-a) concentrations from water‐leaving radiance. NASA’s standard approach since the SeaWiFS era has been the family of “OCx” band‐ratio algorithms, which use blue-to-green reflectance ratios fit to in situ Chl data via polynomial regression [@OReilly2000]. These O’Reilly et al. formulations (e.g., OC4 for SeaWiFS, OC3 for MODIS) became the default global chlorophyll algorithms for multiple sensors due to their relative simplicity and consistency in Case-1 waters. In 2014, NASA introduced the Ocean Color Index (OCI) algorithm, a hybrid empirical model that blends a three‐band reflectance difference method for low‐Chl regimes with the traditional OCx ratio approach at higher Chl concentrations [@NASA2014; @NASA_OC4]. This transition from pure band ratios to a band-difference (baseline subtraction) technique was motivated by the need for improved cross-sensor consistency and better performance in oligotrophic waters. Notably, the color‐index method (CI; Hu et al., 2012) was found to be far less sensitive to instrument noise and atmospheric correction errors than the band‐ratio algorithms. As a result, the OCI approach yields smoother, more stable chlorophyll retrievals in low‐Chl oceans, reducing spurious spatial variability and achieving closer agreement between sensors (SeaWiFS, MODIS, VIIRS) [@hu2019improving].

#### Limitations of Existing Approaches
Despite their empirical success, the existing chlorophyll algorithms have fundamental statistical limitations. The regression procedure underpinning OCx algorithms effectively optimizes the relationship in a forward sense – maximizing the likelihood of the observed reflectance given a chlorophyll value $p(Rrs | Chl)$ – rather than solving the inverse problem of interest $p(Chl | Rrs)$. In practice, the coefficients are tuned to reproduce the sample data distribution, which can lead to overfitting the calibration dataset and biased retrievals when applied broadly. Moreover, uncertainty quantification in these algorithms is ad hoc at best; they provide a point estimate of Chl-a with empirically derived error bounds, but no rigorous probabilistic measure of confidence for each prediction. Another drawback is the use of high-order polynomial fits in the OCx formulations that, while flexible, reduce the interpretability of the model’s spectral response. For example, the standard OC4 algorithm expresses log-transformed Chl-a as a fourth-degree polynomial function of a blue/green reflectance ratio [@OReilly2000]. Such complex polynomial relationships make it difficult to discern how any single band influences the chlorophyll estimate, obscuring physical insights. In summary, current empirical algorithms are not formulated for optimal inference and transparency: they are calibrated to mimic data rather than to robustly infer chlorophyll from reflectance, and they lack a formal mechanism to provide per-observation uncertainty or mechanistic interpretation of the spectral inputs.

>>>>

---
title: "Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation "
author:  
  - name: Erdem M. Karaköylü
    orcid: 0000-0002-6156-1720
    corresponding: true
    email: erdemk@protonmail.com
    roles:
      - Investigation
      - Software
      - Visualization
    affiliations:
      - Consultant
  - name: Susanne E. Craig
    orcid: 0000-0002-8963-0951
    corresponding: true
    email: susanne.e.craig@nasa.gov
    roles:
      - Investigation
      - Project administration
    affiliations:
      - NASA
abstract: Placeholder
date: last-modified
bibliography: references.bib
---

@@

#### Historical Context of Chlorophyll Algorithms
Satellite ocean color observations have long been fundamental for monitoring marine ecosystems, as they enable the global estimation of chlorophyll‑a—a key indicator of phytoplankton biomass and ocean productivity. Early empirical algorithms, notably developed by O’Reilly et al. [@oreilly1998; @oreilly2000], established the $OCx$ family (where $x$ denotes the number of bands used) of polynomial regression models. These models relate blue-to-green reflectance ratios (after log‑transformation) to in situ chlorophyll‑a, employing either straight band ratios (BR) or maximum band ratios (MBR)—the latter selecting the highest available blue-to-green ratio for any given observation as input to a high‑order polynomial. These formulations have served as the operational foundation for chlorophyll‑a products across a broad range of satellite ocean color sensors—from the pioneering Coastal Zone Color Scanner (CZCS) through SeaWiFS, MODIS, and MERIS to more recent missions—offering a straightforward and robust approach for Case‑1 waters. However, their performance is more limited in optically complex Case‑2 waters and remains sensitive to atmospheric correction errors.

Subsequent refinements were introduced to address these deficiencies. For example, Hu et al. [@hu2012novel] proposed a Color Index (CI) formulation that employs a band‑difference approach to reduce sensitivity to residual atmospheric errors and instrument noise, with further improvements enhancing inter‑sensor agreement [@hu2019improving]. The increasing availability of calibration data (e.g., [@Valente2015]) and ongoing algorithmic improvements have led to the development of additional variants of the $OCx$ algorithms—specifically, the OC5 and OC6 formulations. O’Reilly and Werdell [@oreilly2019] maintain that OC5 extends the spectral basis by incorporating the 412 nm band, thereby exploiting its strong signal in clear, oligotrophic waters, while OC6 replaces the traditional denominator with the mean of the 555 and 670 nm reflectances, improving the dynamic range at low chlorophyll concentrations. In total, [@oreilly2019] propose 65 versions of BR/MBR $OCx$ type algorithms for 25 sensors—on average, two or more variants per sensor. With this arsenal, it is hoped, researchers are better equipped to address the wide array of bio‑optical environments encountered in global ocean color applications.

#### Limitations of Existing Approaches
Unfortunately, the development process for traditional ocean color algorithms is founded on a statistical fallacy. Frequentism—the classical approach—assumes that observations are random samples drawn from the studied phenomena. These samples' probabilities are  considered the observed frequencies that capture the the true - albeit noised up and potentially biased - behavior of the system. Under this framework, if a phenomenon is hypothesized to be represented by a model $M$ and data $$D$ is collected, then the model’s representativeness is evaluated solely through the likelihood $p(D∣M)$, which measures the probability of observing the data given that $$M$ is true. The fallacy arises from the mistaken belief that, with sufficiently large amounts of data, one can simply "flip" the likelihood to obtain $p(M∣D)$, thereby yielding a probabilistic statement about $M$. Compounding this error is the imposition of numerous unverified assumptions—disguised as objectivity—that lead frequentists to set the parameters of $M$ by maximizing $p(D∣M)$, under the false premise that this provides the best possible representation of the model. In essence, this fallacy leads to a confusion between the sampling probability $p(D∣M)$ and the inferential probability $p(M∣D)$; the latter  when correctly obtained  quantifies the plausibility of the hypothesis represented by $M$ - precisely what we seek. The historical dimensions of this frequentist illusion are fascinating in their own right, and we invite interested readers to consult works such as McGrayne [@mcgrayne] and Clayton [@clayton].

Bayesianism, Frequentism's archrival, provides the framoework to establish the plausibility of $M$ using observations through the reinterpretation of how probability should be construed. As elaborated extensively by Jaynes [@jaynes2003probability], and more recently by practitioners like Gelman [@gelman2013bayesian], and McElreath [@mcelreath2015statistical], probability is best understood as a statement of our knowledge rather than as an inherent property of the data. the Bayesian perspective treats the data, once observed, as fixed and the model parameters as latent random variables the uncertainty of which  can—and should—be explicitly modeled via prior distributions. Prior distributions are probabilistic constructs that help workers encode prior knowledge. Priors can represents knowledge elicited from subject matter experts, assumptions held by the modeler or a complete absence thereof.

This conceptual framework dissolves the false dichotomy between empirical and semi‑analytical approaches. Both are, in essence, different hypotheses or approximations about how ocean color data are generated; they merely provide alternative expressions of the likelihood 
$p(Chl | Rrs)$ to be updated in light of new data. By adopting a Bayesian paradigm, our approach integrates prior knowledge with observed data to yield a full posterior predictive distribution—offering not only improved predictive performance but also principled uncertainty quantification. In doing so, it circumvents the need for the proliferation of narrowly tuned, sensor-specific variants (65 versions for 25 sensors, as documented by O’Reilly and Werdell [@OREILLY2019]) that are symptomatic of ad hoc calibration processes. Detailed methodological comparisons and additional model refinements are provided in the Supplementary Material.
